---
title: "HW3_yc4384"
author: "Yangyang Chen"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret) 
library(tidymodels) 
library(splines)
library(mgcv)
library(pdp)
library(earth) 
library(tidyverse) 
library(ggplot2)
library(viridis)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

## Problem_1

In this exercise, we explore the application of nonlinear models to analyze the “College” dataset, comprising statistics from 565 US colleges as reported in a past issue of US News and World Report. The response variable is the out-of-state tuition (Outstate). 

### Data Importing and Cleaning


```{r}
college_df = 
  read.csv("College.csv", header = TRUE) |> 
  janitor::clean_names() |> 
  mutate(
    out_state = outstate
  ) |> 
  select(-outstate)
colnames(college_df)
# matrix of predictors
x = model.matrix(out_state ~ ., college_df)[, -1] # vector of response
y = college_df$out_state
```

Find Nonlinear Variables:
```{r}
scatterplots <- lapply(names(college_df)[-c(1,18)], function(var) {
  ggplot(data = college_df, aes_string(x = var, y = "out_state")) +
  geom_point(color="darkgreen", size = 0.1, alpha = 0.3) +  
  theme_minimal() +  
  labs(title = paste("outstate vs.", var),
       x = var, y = "Outstate") + 
  theme(axis.text = element_text(size = 5),  # Adjust font size
        legend.position = "none") 
})

gridExtra::grid.arrange(grobs = scatterplots, ncol = 4)
```
We use scatterplot to explore the relationship between the `out_state` and other variables. The variable `books`, `perc.alumni` and `s_f_ratio` shows potentially nonlinear trend.

### (a) Fit smoothing spline models to predict out-of-state tuition (Outstate) using the percentage of alumni who donate (perc.alumni) as the only predictor, across a range of degrees of freedom. Plot the model fits for each degree of freedom.

```{r}
perc.grid = seq(from = -10, to = 50, by = 1)
fit.ss.percs =
  perc.grid |> 
  lapply(function(df) smooth.spline(college_df$perc_alumni, college_df$out_state, df = perc.grid))

# 
# fit.ss.perc = 
#   smooth.spline(college_df$perc, college_df$out_state) 
# fit.ss.perc$df
# fit.ss.perc$gcv
```

```{r}
plot_data = 
  data.frame(x = college_df$perc_alumni) |> 
  mutate(
    across(everything(), ~ predict(fit.ss.percs[[1]], .)$y)
    )

names(plot_data)[-1] = paste0("y_df_", perc.grid)

plot_data = reshape2::melt(plot_data, id.vars = "x", variable.name = "degree_of_freedom", value.name = "y")

ggplot(plot_data, aes(x = x, y = y, color = degree_of_freedom)) +
  geom_point(data = college_df, aes(x = perc_alumni, y = out_state), alpha = 0.6) +
  geom_line(aes(group = degree_of_freedom), size = 1) +
  labs(title = "Smoothing Spline Fits Across Degrees of Freedom",
       x = "Percentage of Alumni Who Donate",
       y = "Out-of-State Tuition") +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal()
# perc.grid = seq(from = -10, to = 110, by = 1)
# pred.ss.perc = predict(fit.ss.perc, x = perc.grid)
# pred.ss.df = data.frame(pred = pred.ss.perc$y, perc_alumni = perc.grid)
# p = ggplot(data = college_df, aes(x = perc_alumni, y = out_state)) + geom_point(color = rgb(.2, .4, .2, .5))
# p+geom_line(aes(x = perc_alumni, y = pred), data = pred.ss.df,
# color = rgb(.8, .1, .1, 1)) + theme_bw()
```

